**The Heart can Lie: The Role of Interoception and Theory of Mind in Deception**

### 2022-11-21 paper/manuscript.Rmd
anonimize data and revise ms

To ensure that our outcome variables were sensitive to the experimental manipulations, we first tested the effect of the question phrasing (direct *vs.* indirect) and condition (polygraph *vs.* interrogation) on the outcome variables. This analysis was performed using mixed models with the participants and questions both entered as random factors. Marginal contrasts analysis was also performed to clarify the differences between conditions. To allow for a better quantification of the uncertainty associated with the effects, as well as increase the robustness to outliers and artifactual findings, all statistics were undertaken under the Bayesian framework [@makowski2019bayestestr], using informative priors centred around 0 ($t_{Confidence}(1, 0, 1)$, $t_{RT}(1, 0, 3)$, $t_{Heartrate}(1, 0, 8)$).

@@

Compared to truth, lies were rated with less confidence ($\beta=-1.09,~95\%~CI~[-0.94, -1.25],~pd = 100\%$). Moreover, lies told in the polygraph condition were rated with less confidence ($\delta=-0.24,~95\%~CI~[-0.10, -0.39],~pd > 99\%$). The RT did not differ between truth and lies, but was significantly slower in the polygraph condition ($\beta=0.14,~95\%~CI~[0.03, 0.25],~pd > 99\%$) for both conditions. The heart rate was significantly more elevated during lies as compared to truth ($\beta=1.31,~95\%~CI~[2.14, 0.50],~pd > 99.9\%$), and during interrogation as compared to the polygraph condition ($\beta=4.99 ,~95\%~CI~[5.82, 4.19],~pd = 100\%$).

The indirect phrasing of the question only had a significant effect on RT ($\beta=0.34,~95\%~CI~[0.23, 0.44],~pd = 100\%$), leading to slower answers, regardless of whether they were lies or truths.

@@

The 8 MAIA dimensions and the 3 HCT components were reduced to 4 factors (explaining 65.17% of variance). The first factor, labelled *Interoception - Meta* (23.59%), was loaded primarily by Attention Regulation (.97), Self-regulation (.63), Emotional awareness (.60), and Noticing (.49) dimensions of the MAIA and the HCT confidence score (.40). The second factor, labelled *Interoception - Listening* (18.54%), was primarily loaded by the Body Listening (.92) and Trusting (.53) MAIA dimensions, and the Awareness (-.60) and Confidence (.46) HCT scores. The third factor, labelled *Interoception - Focus* (12.07%), was primarily loaded by MAIA Not-Distracting (.87), Emotional Awareness (-.40) and HCT Accuracy (.33). The fourth factor, labelled *Interoception - Regulation* (10.97%), was primarily loaded by MAIA not-worrying (.71), HCT Accuracy (.61) and MAIA Trusting (.40). However, due to the interpretation difficulty, we will refer to these dimensions with their main correlates in **Figure 1** [**Note to fix caption for Figure 1 from HCT Trusting to MAIA trusting**].

@@

The composite *TOM* score was significantly associated with less confident ($\beta=-0.19,~95\%~CI~[-0.36, -0.02],~pd = 98.47\%$) and slower lies ($\beta=0.42,~95\%~CI~[0.01, 0.83],~pd = 97.67%\%$), specifically in the polygraph condition. No significant effect was found with regards to heart rate, and no correlation was observed with dispositional lying traits.

@@

The *Regulation* interoception score was significantly associated with an increased confidence in lies in the polygraph ($\beta=0.32,~95\%~CI~[0.14, 0.51],~pd = 99.99\%$) and the interrogation conditions ($\beta=0.18,~95\%~CI~[0.00, 0.36],~pd = 97.42\%$), and with a decreased confidence in truth only in the polygraph condition ($\beta=-0.1,~95\%~CI~[-0.36, 0.01],~pd = 97.16\%$). No significant association was found with RT or heart rate.

@@

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions ("Polygraph" and "Interrogation") hypothesizingly enhancing each of the two mechanisms. Interestingly, we found a significant difference in participants' lying ability (as measured by response time, heart rate and confidence ratings) between the *Polygraph* and *Interrogation* conditions. This hence confers support for the manipulation used in our experimental paradigm, demonstrating that increasing the attentional focus towards visual interoceptive cues, even if they are fake, can elicit interoceptive awareness and enhance the precision of internal signals if such external cues of interoceptive signals are perceived to be real. Extending previous research pertaining to the application of interoceptive training for improving emotion regulation [@sugawara2020effects; @quadt2021interoceptive; @wu2022enhanced], our study thus provides evidence that interoceptive awareness can be manipulated and thereby modified to guide our emotional responses.

<!-- Interoceptive vs Social Cues -->
Using a bi-modal feedback deception task, we found partial support for our hypothesis. Specifically, participants reported lower lie confidence and took a longer time to respond, but were less physiologically aroused in the *Polygraph* condition, relative to the *Interrogation* condition. While this mixed finding is challenging to interpret, it is possible that one's interoceptive sensitivity, enhanced by the perceived interoceptive cues, interacted with the stress experienced in such deceptive interactions and led to a depletion of cognitive resources [@ten2015physically; @porter2012secrets]. Indeed, the experience of excessive stress has been shown to impair interoceptive accuracy, leading to a hyper-focus on internal bodily signals [@schulz2015interoception]. Furthermore, this is also in line with the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available for the actual act of telling a lie, thereby resulting in individuals taking a longer time to respond, and rating their lies with less confidence [@arnold2019interoception; @hanania2010selective]. This interplay of interoception and maladaptive stress response, and their effect on lying ability hence underscores the complexity underlying successful deceptive behaviour.

<!-- Meta-Interoception, Emotion Regulation and Lying -->
Consistent with our hypothesis, a significant positive association was predominantly found between metacognitive dimensions of interoception (*Meta* and *Regulation* interoception) and individuals' lie confidence when their physiological cues were salient. In contrast to previous studies [@pinna2020systematic; @fustos2013embodiment; @owens2018investigating; @pollatos2007heart], we did not find any significant relationship between individuals' interoception scores and objective heart rate. Taken together, these findings suggest that interoceptive awareness, in and of itself, is insufficient for adaptive emotion regulation. Indeed, some researchers have found positive links between interoceptive awareness and individuals' use of maladaptive strategies such as rumination [@tan2023interoceptive]. Furthermore, hypervigilance to bodily signals is also positively related to somatization and psychopathological disorders such as anxiety [@wang2020intrinsic; @dunn2010can; @pollatos2007interoceptive]. As such, further studies are necessary to examine what factors modulate the relationship between the metacognitve facets of interoception and emotion regulation.

<!-- TOM -->
Our results show a significant association between individuals' composite TOM score with lower lie confidence and longer response time when interoceptive cues were highly salient. Under the free energy principle [@friston2010free], interoception is regarded as integral in TOM [@shah2017heart; @palmer2018going; @ondobaka2017role]. Specifically, during social inference, interoceptive signals are integrated with exteroceptive (e.g., visual and auditory) and proprioceptive information (i.e., motor) to guide one's social behaviour [@ondobaka2017role]. Given recent work in which TOM is associated with one having more precise interoceptive states [@tsakiris2017multisensory; @palmer2018going; @arnold2019interoception], it is possible that individuals with greater TOM abilities have higher interoceptive awareness, more stable self-other representations and less malleable percepts of self [@tsakiris2017multisensory; @arnold2019interoception]. As such, external cues of their interoceptive signals may be perceived as incongruent, hence generating a sense of uncertainty in their responses and leading to their lower lie confidence and longer response times.

<!-- Limitations and Future Research -->
Despite longstanding research interest in deception, little attention has been conferred to understanding the inter-individual differences in lying ability. In this study, we found a significant effect of interoception, particularly pertaining to its metacognitive dimensions, on one's ability to lie convincingly. However, future research is needed to address a few limitations. Namely, objective interoceptive accuracy was only assessed using the HCT. While the HCT is one of the most commonly used measure of objective interoceptive accuracy [@desmedt2022does], concerns regarding its validity has been highlighted in several studies [@desmedt2018heartbeat; @desmedt2022does; @brener2016towards; @legrand2022heart]. Given increasing research efforts invested in the development of novel interoception tasks [@plans2021measuring; @legrand2022heart; @ponzo2021measuring], future works should further examine the relationship between interoception and lying ability using measures with better psychometric properties. Additionally, in order to account for the small sample size, we used factor analysis to reduce the number of dimensions assessing our variables of interest, which could have over-simplified the effect of interoception observed. Taking into perspective that the various facets of interoception are widely regarded as dissociable [@suksasilp2022towards; @mehling2012multidimensional], future studies could thus investigate their effects separately, to better delineate how interoceptive processes influence lying ability.

@@

We would like to thank Jia Rong **what's his full name?** for his help in data collection and **...** for data analysis.

>>>>

To ensure that our outcome variables were sensitive to the experimental manipulations, we first tested the effect of the question phrasing (direct *vs.* indirect) and condition (polygraph *vs.* interrogation) on the outcome variables. This analysis was performed using mixed models with the participants and questions both entered as random factors. Marginal contrasts analysis (denoted by $\Delta$) was also performed to clarify the differences between conditions. To allow for a better quantification of the uncertainty associated with the effects, as well as increase the robustness to outliers and artifactual findings, all statistics were undertaken under the Bayesian framework [@makowski2019bayestestr], using informative priors centred around 0 ($t_{Confidence}(1, 0, 1)$, $t_{RT}(1, 0, 3)$, $t_{Heartrate}(1, 0, 8)$).

@@

Compared to truth, lies were rated with less confidence ($\Delta=-1.35,~95\%~CI~[-1.46, -1.23],~pd = 100\%$), but no difference between the conditions was found. On the other hand, the RT did not differ between truth and lies, but was significantly slower in the polygraph condition for both conditions ($\Delta=0.25,~95\%~CI~[0.62, 0.41],~pd = 100\%$). The heart rate was significantly more elevated during lies as compared to truth ($\Delta=1.16,~95\%~CI~[0.57, 1.73],~pd = 100\%$), and during interrogation as compared to the polygraph condition ($\Delta=4.84 ,~95\%~CI~[4.23, 5.44],~pd = 100\%$).

The indirect phrasing of the question only had a significant effect on RT ($\beta=0.36,~95\%~CI~[0.21, 0.51],~pd = 100\%$), leading to slower answers, regardless of whether they were lies or truths.

@@

The 8 MAIA dimensions and the 3 HCT components were reduced to 4 factors (explaining 65.17% of variance). The first factor, labelled *Interoception - Meta* (23.59%), was loaded primarily by Attention Regulation (.97), Self-regulation (.63), Emotional awareness (.60), and Noticing (.49) dimensions of the MAIA and the HCT confidence score (.40). The second factor, labelled *Interoception - Listening* (18.54%), was primarily loaded by the Body Listening (.92) and Trusting (.53) MAIA dimensions, and the Awareness (-.60) and Confidence (.46) HCT scores. The third factor, labelled *Interoception - Focus* (12.07%), was primarily loaded by MAIA Not-Distracting (.87), Emotional Awareness (-.40) and HCT Accuracy (.33). The fourth factor, labelled *Interoception - Regulation* (10.97%), was primarily loaded by MAIA not-worrying (.71), HCT Accuracy (.61) and MAIA Trusting (.40). 
<!-- However, due to the interpretation difficulty, we will refer to these dimensions with their main correlates in **Figure 1** [**Note to fix caption for Figure 1 from HCT Trusting to MAIA trusting**]. -->

@@

The composite *TOM* score was significantly associated with less confident ($\beta=-0.19,~95\%~CI~[-0.36, -0.02],~pd = 98.47\%$) and slower lies ($\beta=0.42,~95\%~CI~[0.01, 0.83],~pd = 97.67%\%$), specifically in the polygraph condition. No significant effect was found with regards to heart rate, and no correlation was observed with dispositional lying traits (**Figure 1**).

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="TODO: Caption."}
knitr::include_graphics("../figures/figure1.png")
```

@@

The *Regulation* interoception score was significantly associated with an increased confidence in lies in the polygraph ($\beta=0.32,~95\%~CI~[0.14, 0.51],~pd = 99.99\%$) and the interrogation conditions ($\beta=0.18,~95\%~CI~[0.00, 0.36],~pd = 97.42\%$), and with a decreased confidence in truth only in the polygraph condition ($\beta=-0.1,~95\%~CI~[-0.36, 0.01],~pd = 97.16\%$). No significant association was found with RT or heart rate (**Figure 2**).

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="TODO: Caption."}
knitr::include_graphics("../figures/figure2.png")
```

@@

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions ("Interrogation" and "Polygraph") designed to enhance each of the two mechanisms. Interestingly, we found that when participants were displayed a (fake) physiological feedback (the polygraph condition), instead of a face of a person they had to lie to (the interrogation condition), their response time for both lies and truths was increased, as well as their heart rate. Although the condition did not impact the subjective confidence that participants had in their answers, it suggests that believing to be submitted to a lying detection machine that supposedly reads physiological signals is a harder and/or more stressful condition than answering to a person.

<!-- This hence confers support for the manipulation used in our experimental paradigm, demonstrating that increasing the attentional focus towards visual interoceptive cues, even if they are fake, can elicit interoceptive awareness and enhance the precision of internal signals if such external cues of interoceptive signals are perceived to be real. Extending previous research pertaining to the application of interoceptive training for improving emotion regulation [@sugawara2020effects; @quadt2021interoceptive; @wu2022enhanced], our study thus provides evidence that interoceptive awareness can be manipulated and thereby modified to guide our emotional responses. -->

<!-- Interoceptive vs Social Cues -->
<!-- Using a bi-modal feedback deception task, we found partial support for our hypothesis. Specifically, participants reported lower lie confidence and took a longer time to respond, but were less physiologically aroused in the *Polygraph* condition, relative to the *Interrogation* condition. While this mixed finding is challenging to interpret, it is possible that one's interoceptive sensitivity, enhanced by the perceived interoceptive cues, interacted with the stress experienced in such deceptive interactions and led to a depletion of cognitive resources [@ten2015physically; @porter2012secrets]. Indeed, the experience of excessive stress has been shown to impair interoceptive accuracy, leading to a hyper-focus on internal bodily signals [@schulz2015interoception]. Furthermore, this is also in line with the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available for the actual act of telling a lie, thereby resulting in individuals taking a longer time to respond, and rating their lies with less confidence [@arnold2019interoception; @hanania2010selective]. This interplay of interoception and maladaptive stress response, and their effect on lying ability hence underscores the complexity underlying successful deceptive behaviour. -->

<!-- TOM -->
Our results suggest that higher ToM abilities were related to a slower and less confident lies, but only in the polygraph condition. One possible interpretation is that people with stronger ToM abilities rely by default more on it when lying (i.e., they focus on - and try to read - the other person). When that mechanism is cut off (i.e., when there is no person to lie to - but in our case a "machine"), their lying ability decreases.

<!-- Interoception -->
We also showed that interoceptive abilities are correlated with a higher confidence in one's lies in the polygraph condition, a condition in which the attention towards internal reactions is fostered. However, in contrast to previous studies [@pinna2020systematic; @fustos2013embodiment; @owens2018investigating; @pollatos2007heart], we did not find any significant relationship between individuals' interoception scores and the heart rate changes during their answers. This points toward a predominantly meta-cognitive effect without necessarily an actual bodily regulation (i.e., participants with good interoception feel that their lies are more convincing, but do not actively attenuate their bodily reactions).

<!-- Our results show a significant association between individuals' composite TOM score with lower lie confidence and longer response time when interoceptive cues were highly salient. Under the free energy principle [@friston2010free], interoception is regarded as integral in TOM [@shah2017heart; @palmer2018going; @ondobaka2017role]. Specifically, during social inference, interoceptive signals are integrated with exteroceptive (e.g., visual and auditory) and proprioceptive information (i.e., motor) to guide one's social behaviour [@ondobaka2017role]. Given recent work in which TOM is associated with one having more precise interoceptive states [@tsakiris2017multisensory; @palmer2018going; @arnold2019interoception], it is possible that individuals with greater TOM abilities have higher interoceptive awareness, more stable self-other representations and less malleable percepts of self [@tsakiris2017multisensory; @arnold2019interoception]. As such, external cues of their interoceptive signals may be perceived as incongruent, hence generating a sense of uncertainty in their responses and leading to their lower lie confidence and longer response times.   -->

<!-- Meta-Interoception, Emotion Regulation and Lying -->
<!-- Consistent with our hypothesis, a significant positive association was predominantly found between metacognitive dimensions of interoception (*Meta* and *Regulation* interoception) and individuals' lie confidence when their physiological cues were salient. In contrast to previous studies [@pinna2020systematic; @fustos2013embodiment; @owens2018investigating; @pollatos2007heart], we did not find any significant relationship between individuals' interoception scores and objective heart rate. Taken together, these findings suggest that interoceptive awareness, in and of itself, is insufficient for adaptive emotion regulation. Indeed, some researchers have found positive links between interoceptive awareness and individuals' use of maladaptive strategies such as rumination [@tan2023interoceptive]. Furthermore, hypervigilance to bodily signals is also positively related to somatization and psychopathological disorders such as anxiety [@wang2020intrinsic; @dunn2010can; @pollatos2007interoceptive]. As such, further studies are necessary to examine what factors modulate the relationship between the metacognitve facets of interoception and emotion regulation. -->

<!-- Limitations and Future Research -->
Although yielding promising results, this exploratory study is limited by the sample size. Although we tried to mitigate it by 1) extracting more robust variables (by combining multiple ones by means of feature reduction) and 2) using a suited analysis approach (Bayesian statistics with informative priors), future replication studies with larger samples are warranted to confirm this first investigation. Nonetheless, we believe our results to be credible as we find consistent patterns across various facets and measures (for instance, all interoceptive dimensions, although distinct, share a similar trend) in line with theoretical expectations. The statistical power could also explain the overall lack of results with heart rate, which has a higher signal-to-noise ratio as compared to subjective reports (such as confidence scales).

Another aspect to note is the strong reliance on self-reported measures as outcome variables of lie ability (in particular, the measure of answer confidence, but also the auto-questionnaires). This might conflate meta-cognitive abilities as well as dishonest answers. Although we tried to include more objective measures, such as RT (although it too was tied to the participants' conscious decision to press a key) and heart rate, future studies should attempt at measuring objectively the answer (lie or truth) quality, for instance by means of external examiners. Note that this is not limitation *per se*, as it answers a slightly different question  - what are the correlates of *objective* lying skills - than deception self-confidence.

Additionally to limitations pertaining to the measure of lying ability, care has to be given to the measure of the predictor constructs, namely ToM and interoception. While we tried to include an behavioral task as well as a subjective questionnaire for each, it has to be underlined that they are notoriously difficult concepts to measure. In particular, objective interoceptive accuracy was assessed using the Heartbeat Counting Task. While the HCT used to be considered as a gold standard and is one of the most commonly used measure [@desmedt2022does], concerns regarding its validity has been highlighted in several studies [@desmedt2018heartbeat; @desmedt2022does; @brener2016towards; @legrand2022heart]. Given increasing research efforts invested in the development of novel interoception tasks [@plans2021measuring; @legrand2022heart; @ponzo2021measuring], future works should further examine the relationship between interoception and lying ability using measures with better psychometric properties. Additionally, our application of feature reduction as a noise-elimination measure could have over-simplified the data. A more complex pattern of relationships, with different contributions of various subdimensions of ToM and interoception, could emerge provided a sufficient statistical power and valid measures.

<!-- Taking into perspective that the various facets of interoception are widely regarded as dissociable [@suksasilp2022towards; @mehling2012multidimensional], future studies could thus investigate their effects separately, to better delineate how interoceptive processes influence lying ability.    -->

In conclusion, this study study is a first step towards assessing the contribution of ToM and interoception abilities in deception in one's ability to lie convincingly. We introduced a new paradigm that is able to modulate the contribution of these mechanisms while remaining relevant to applied fields of lie detection and criminology (in which the experimental conditions find echoing practices). Our results provide princeps evidence that interoception could be a key - and overlooked - mechanism in deception.

@@

We would like to thank Jia Rong **what's his full name?**, Max (**full name**, partial RA) for his help in data collection and **...** for data analysis.
